---
title: "Breast Cancer Diagnostics Analysis"
author: "Etini Levi"
date: "2026-04-15"
output:
  html_document: default
---

```{r setup, include=TRUE, echo = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(janitor)
UCI_Breast_Cancer_Dataset <- read_csv("C:/Users/hp/Desktop/Data Analytics and Science/Datasets/UCI Breast Cancer Dataset.csv")

df <- UCI_Breast_Cancer_Dataset
my_data <- df %>%
  clean_names() %>%
  select(-where(~all(is.na(.)))) 
```

## Brief Summary

This report analyzes the Wisconsin Breast Cancer (Diagnostic) dataset from <https://www.kaggle.com/datasets/sarahvch/breast-cancer-wisconsin-prognostic-data-set> to determine the relationship between cellular morphology and malignancy. 

Using Logistic Regression, we achieved a classification accuracy of 92%.

We begin by examining the distribution of cell nuclei radius. 

Larger nuclei are statistically more likely to be associated with malignant (M) samples.

```{r df, echo=FALSE}
ggplot(df, aes(x = diagnosis, y = radius_mean, fill = diagnosis)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Mean Radius: Malignant vs. Benign",
       x = "Diagnosis (M = Malignant, B = Benign)",
       y = "Mean Radius of Nuclei")
```

## Principal Component Analysis
The dataset had more then 30 features, so a PCA was performed to reduce the dataset in 2D space; PC1 and PC2
There was a clear separation between the two groups suggesting a strong biological signal.
```{r pca_df, echo=FALSE}
numeric_df <- df %>% select(where(is.numeric))
pca_res <- prcomp(numeric_df, scale. = TRUE)
pca_df <- as.data.frame(pca_res$x)
pca_df$diagnosis <- df$diagnosis

ggplot(pca_df, aes(x = PC1, y = PC2, color = diagnosis)) +
  geom_point(alpha = 0.6, size = 3) +
  theme_classic() +
  labs(title = "PCA: Visualizing Group Separation",
       x = "Principal Component 1",
       y = "Principal Component 2")
```


## Predictive Modeling
  A Logistic Regression model was used to perform the predictive modeling. 
The focus were on three key features: Mean Radius, Mean Texture, and Mean Smoothness.
The model was evaluated on a held-out test set (25% of the data) and the accuracy score was 92.9%
The confusion matrix showed:
83 True Negative and 4 False Negative
6 False Positive and 49 True Positive.

```{r, echo=FALSE}
library(rsample)

df$diagnosis <- as.factor(df$diagnosis)
set.seed(123)
data_split <- initial_split(df, prop = 0.75, strata = diagnosis)
train_data <- training(data_split)
test_data  <- testing(data_split)

model <- glm(diagnosis ~ radius_mean + texture_mean + smoothness_mean, 
             data = train_data, 
             family = "binomial")

summary(model)

test_data$prob <- predict(model, newdata = test_data, type = "response")


test_data$prediction <- ifelse(test_data$prob > 0.5, "M", "B")


table(Predicted = test_data$prediction, Actual = test_data$diagnosis)
```

## Conclusion
From this project, we have been able to perform EDA with the dataset, performed PCA to reduce 34 features into 2D, and then developed a Logistic Regression model to perform a predictive analytics of diagnosis based on radius_mean, texture_mean and smoothness_mean. The model had a 92.9% accuracy.
